Modal Abstractions for Virtualizing Memory Addresses

### 2025-03-24 examples.tex
finished advisor pass, switched to author-year citations, now 23 pages with a little wiggle room
fixes

\begin{figure}\footnotesize
  \vspace{-1em}
$\begin{array}{l}\left( \begin{array}{l} \ulcorner \textsf{pdp}+\textsf{PDPEX(va)} = \textsf{pdp\_entry} \land \textsf{table\_root(pml4\_entry}_{\textsf{pfn}})  = \textsf{pdp} \\ \land \textsf{pd}+\textsf{PDEX(va)} = \textsf{pd\_entry} \land \textsf{table\_root(pml3\_entry}_{\textsf{pfn}})  = \textsf{pd} \\ \land \textsf{pt}+\textsf{PDPEX(va)} = \textsf{pt\_entry} \land \textsf{table\_root(pml2\_entry}_{\textsf{pfn}})  = \textsf{pt} \urcorner  \ast \\  \textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus\{\textsf{pml4\_entry},\textsf{pdp\_entry},\textsf{pd\_entry},\textsf{pt\_entry}\},m)  \\ \ghostmaptoken{\delta{}s}{\rtv}{\delta} \ast  \\
      \ghostmaptoken{\textsf{id}}{(\mathsf{pml4\_entry})}{\textsf{4}}\ast \ghostmaptoken{\textsf{id}}{(\mathsf{plm3\_entry})}{\textsf{3}} \ast \ghostmaptoken{\textsf{id}}{(\mathsf{plm2\_entry})}{\textsf{2}}  \ast \\  \textsf{pml4\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l4e\_val} \ast  \tcbhighmath[boxrule=1pt,drop fuzzy shadow=blue]{\ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present l4e\_val})\urcorner}\;\; \ast \\  \textsf{pdp\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l3e\_val} \ast  \tcbhighmath[boxrule=1pt,drop fuzzy shadow=blue]{\ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present l3e\_val})\urcorner} \;\; \ast \\  \textsf{pd\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l2e\_val} \ast \tcbhighmath[boxrule=1pt,drop fuzzy shadow=blue]{\ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present l2e\_val})\urcorner }  \end{array} \right)  =\textsf{R}_{\textsf{walk?} } \\ \ast \left(\begin{array}{l} \textsf{rax} \mapsto \textsf{pte} \ast  \textsf{pte} \mapsto_{\textsf{vpte,qfrac}} \textsf{pte\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present pte\_val})\urcorner \end{array}\right) =\textsf{R}_{\textsf{l1e}}\end{array} $
\caption{\textsf{R}$_{\textsf{walk?}}$: Resources needed for walking page-table directory via calls to \textsf{pte\_get\_next\_table} in Figure \ref{fig:calltopteinitialize}}
\label{walkpgdir}
\vspace{-1em}
\end{figure}

Realizing a software page-table walk amounts to calling \textsf{pte\_get\_next\_table} for each level as shown in Figure \ref{walkpgdir}. 
The special part of the specficication for a page table walk can be considered as accumulation of memory mappings for the page-table entries 
visited and frame addresses for page-tables. 
For example, Lines \ref{line:ex_l4_vpte} and \ref{line:ex_l3_vpte} in Figure \ref{walkpgdir} show the virtual pte-pointsto assertions for L4 and L3 entries.
In the final post-condition, we expect the accumulation of these resources from each level -- $\textsf{R}_{\textsf{walk}}$ -- 
which allows us to construct and return the path to the L1 entry in the tree to insert a new page.

@@

\subsection{Mapping a New Page}
\label{sec:mapnew}
One of the key tasks of a page fault handler in a general-purpose OS kernel is
to map new pages into an address space by writing into an existing page table via a call
\[\textsf{vaspace\_mappage(pte\_t *pml4, void *va,uintptr\_t fpaddr, uint64\_t perm )}\] in Figure \ref{fig:mapping_code}.
To do so, with a given allocated a fresh page (\textsf{fpaddr}), then calculate the appropriate
known-valid page table walks (via \textsf{walkpgdir} Line \ref{line:call_walkpgdir} in Figure \ref{fig:mapping_code})  and update 
the appropriate L1 page table entry (Line 11 in Figure \ref{fig:mapping_code});
unmapping is the reverse of the logic we discuss here.
\looseness=-1
%\lstset{
%  columns=fullflexible,
%  numbers=left,
%  basicstyle=\ttfamily,
%  keywordstyle=\color{blue}\bfseries,
%  morekeywords={mov,add,call},
%  emph={rsp,rdx,rax,rbx,rbp,rsi,rdi,rcx,r8,r9,r10,r11,r12,r13,r14,r15},
%  emphstyle=\color{green},
%  emph={[2]cr3},
%  emphstyle={[2]\color{violet}},
%  morecomment=[l]{;;},
%  mathescape
%}
\begin{figure}\footnotesize
  \begin{lstlisting}[mathescape,escapeinside={(*}{*)}]
;;complstatus_t vaspace_mappage(pte_t *pml4, void *va,uintptr_t fpaddr, uint64_t perm ) {
... ;;setting up the stack      
 $\specline{\textsf{P} \ast \textsf{pml4} \mapsto_{\textsf{id}} \textsf{\_} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta} \ast \ulcorner\theta \; !!\; \vaddr = \texttt{None}\urcorner}_{\rtv}$      
;; pte_t *pte = walkpgdir(pml4, a);
 mov  -0x8[rbp], rdi;
 mov  -0x16[rbp], rsi;
 mov  -0x24[rbp], rdx;
% mov  rdx, -0x16[rbp];
 mov  rdx, (l4 + PML4EX(va) + natToWord 64 KERNBASE)))
 mov  rsi,  -0x32[rbp];
 mov  rax, 0x0;
 callq  291 <walkpgdir> (*\label{line:call_walkpgdir}*)
 $\specline{ \textsf{R}_{\textsf{walk}} \ast \left(\begin{array}{l}   \textsf{rax} \mapsto \textsf{pte} \ast  \textsf{pte} \mapsto_{\textsf{vpte,qfrac}} \textsf{pte\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present pte\_val}) \end{array}\right) =\textsf{R}_{\textsf{l1e}} }$
;;if(pte->present != 0) {
...        
mov    0x0,r8; mov    rax,rcx; mov    rcx,-0x30[rbp]
... ;; Copy user access bit, writethrough, etc. (not part of our semantics) into pte from perm
;;pte->pfn          = PTE_ADDR_TO_PFN(fpaddr);
mov    -0x18[rbp],rcx; shr    0xc,rcx; mov    -0x30[rbp],rdx; mov    [rdx],rsi ;; Load current entry in rsi
$\specline{\textsf{pt\_entry}_{\textsf{pfn}} \mapsto_{p} \textsf{addr\_to\_pfn}(\textsf{pa}) \ast \textsf{R}_{\textsf{walk\_acc}}}_{\rtv} \Rrightarrow \specline{\textsf{va} \mapsto_{\textsf{v,q}} \textsf{pa} \ast \textsf{P}}$
movabs 0xffffffffff,rdi; and    rdi,rcx; shl    0xc,rcx; movabs 0xfff0000000000fff,rdi
and    rdi,rsi; or     rcx,rsi; mov    rsi,[rdx] ;; Store updated entry back to L1 entry (*\label{line:l1entry_store}*)
;;}
... ;; set present bit in entry, then clean up the stack and return (proof re-packs invariants)
;;}
\end{lstlisting}
\vspace{-1em}
  \caption{Specification and proof of code for mapping a new page with $\textsf{R}_{\textsf{walk\_acc}}$ in Figure \ref{walkpgdir} expanded.}
\label{fig:mapping_code}
\end{figure}

>>>>

\begin{figure}\footnotesize
  \vspace{-1em}
$\begin{array}{l}\left( \begin{array}{l} \ulcorner \textsf{pdp}+\textsf{PDPEX(va)} = \textsf{pdp\_entry} \land \textsf{table\_root(pml4\_entry}_{\textsf{pfn}})  = \textsf{pdp} \\ \land \textsf{pd}+\textsf{PDEX(va)} = \textsf{pd\_entry} \land \textsf{table\_root(pml3\_entry}_{\textsf{pfn}})  = \textsf{pd} \\ \land \textsf{pt}+\textsf{PDPEX(va)} = \textsf{pt\_entry} \land \textsf{table\_root(pml2\_entry}_{\textsf{pfn}})  = \textsf{pt} \urcorner  \ast \\  \textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus\{\textsf{pml4\_entry},\textsf{pdp\_entry},\textsf{pd\_entry},\textsf{pte}\},m)  \\ \ghostmaptoken{\delta{}s}{\rtv}{\delta} \ast  \\
      \ghostmaptoken{\textsf{id}}{(\mathsf{pml4\_entry})}{\textsf{4}}\ast \ghostmaptoken{\textsf{id}}{(\mathsf{plm3\_entry})}{\textsf{3}} \ast \ghostmaptoken{\textsf{id}}{(\mathsf{plm2\_entry})}{\textsf{2}}  \ast \\  \textsf{pml4\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l4e\_val}  \ast \\  \textsf{pdp\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l3e\_val}  \ast \\  \textsf{pd\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l2e\_val}  \end{array} \right)  =\textsf{R}_{\textsf{walk} } \\ \ast \left(\begin{array}{l} \textsf{rax} \mapsto \textsf{pte} \ast  \textsf{pt\_entry} \mapsto_{\textsf{vpte,qfrac}} \textsf{pte} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present pte\_val})\urcorner \end{array}\right) =\textsf{R}_{\textsf{l1e}}\end{array} $
\caption{\textsf{R}$_{\textsf{walk}}$: Resources obtained from invariant for walking page-table directory with \emph{non-existing entries} via calls to \textsf{pte\_get\_next\_table} in Figure \ref{fig:calltopteinitialize}}
\label{walkpgdir}
\vspace{-1em}
\end{figure}
Realizing a software page-table walk amounts to calling \textsf{pte\_get\_next\_table} for each level as shown in Figure \ref{walkpgdir}. 
The special part of the specficication for a page table walk can be considered as accumulation of memory mappings for the page-table entries 
visited and frame addresses for page-tables. 
For example, Lines \ref{line:ex_l4_vpte} and \ref{line:ex_l3_vpte} in Figure \ref{walkpgdir} show the virtual pte-pointsto assertions for L4 and L3 entries.
In the final post-condition, we expect the accumulation of these resources from each level -- $\textsf{R}_{\textsf{walk}}$ -- 
which allows us to construct and return the path to the L1 entry in the tree to insert a new page.

@@

\subsection{Mapping a New Page}
\label{sec:mapnew}
One of the key tasks of a page fault handler in a general-purpose OS kernel is
to map new pages into an address space by writing into an existing page table via a call
\[\textsf{vaspace\_mappage(pte\_t *pml4, void *va,uintptr\_t fpaddr, uint64\_t perm )}\] in Figure \ref{fig:mapping_code}.
To do so, with a given allocated a fresh page (\textsf{fpaddr}), then calculate the appropriate
known-valid page table walks (via \textsf{walkpgdir} Line \ref{line:call_walkpgdir} in Figure \ref{fig:mapping_code})  and update 
the appropriate L1 page table entry (Line 11 in Figure \ref{fig:mapping_code});
unmapping is the reverse of the logic we discuss here.
\looseness=-1
%\lstset{
%  columns=fullflexible,
%  numbers=left,
%  basicstyle=\ttfamily,
%  keywordstyle=\color{blue}\bfseries,
%  morekeywords={mov,add,call},
%  emph={rsp,rdx,rax,rbx,rbp,rsi,rdi,rcx,r8,r9,r10,r11,r12,r13,r14,r15},
%  emphstyle=\color{green},
%  emph={[2]cr3},
%  emphstyle={[2]\color{violet}},
%  morecomment=[l]{;;},
%  mathescape
%}
\begin{figure}\footnotesize
  \begin{lstlisting}[mathescape,escapeinside={(*}{*)}]
;;complstatus_t vaspace_mappage(pte_t *pml4, void *va,uintptr_t fpaddr ) {
... ;;setting up the stack      
 $\specline{\textsf{P} \ast \textsf{pml4} \mapsto_{\textsf{id}} \textsf{\_} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta} \ast \ulcorner\theta \; !!\; \vaddr = \texttt{None}\urcorner}_{\rtv}$      
;; pte_t *pte = walkpgdir(pml4, a);
 mov  -0x8[rbp], rdi;
 mov  -0x16[rbp], rsi;
 mov  -0x24[rbp], rdx;
;; pte = walkpgdir_succ(pml4, va);
 mov    -0x8[rbp], %rdi;
 mov    -0x16[rbp], rsi;
 mov    0x0, rax;
 callq  291 <walkpgdir> (*\label{line:call_walkpgdir}*)
 $\specline{ \textsf{R}_{\textsf{walk}} \ast \left(\begin{array}{l}   \textsf{rax} \mapsto \textsf{pte} \ast  \textsf{pte} \mapsto_{\textsf{vpte,qfrac}} \textsf{pte\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present pte\_val}) \end{array}\right) =\textsf{R}_{\textsf{l1e}} }$
 mov  rax, rdx;
 mov    rdx,-0x28[rbp];
;;if (!pte->present){ 
 mov    -0x28[rbp], rdx
 mov    [rdx], rcx
 and    0x1, rcx
 mov    rcx, r8
 cmp    0x0, r8
 jne    3e0 <vaspace_mappage+0x190>
 ...
;;pte->pfn          = PTE_ADDR_TO_PFN(pa);
 mov    -0x18[rbp], rdx;
 shr    0xc, rdx;
 mov    -0x28[rbp], rsi;
 mov    [rsi], rdi;
 movabs 0xffffffffff, rcx
 and    rcx, rdx
 shl    0xc, rdx
 movabs 0xfff0000000000fff, rcx
 and    rcx, rdi
 or     rdx, rdi
 mov    rdi, [rsi]
$\specline{ \textsf{pte} \mapsto_{\textsf{vpte,1}} (\textsf{set\_pfn}(\textsf{pte\_val}, \textsf{PTE\_ADDR\_TO\_PFN}(\textsf{fpaddr}))) \ast  \ulcorner \lnot(\textsf{entry\_present pte\_val}) \urcorner \ast P}$
 ;; Store updated entry back to L1 entry (*\label{line:l1entry_store}*)
;;}
... ;;set present bit in entry, then clean up the stack
... ;;and return either unaccessible pte pointsto or one with fresh page (fpaddr)
 $\specline{ \begin{array}{l}\textsf{pte} \mapsto_{\textsf{vpte,qfrac}} \textsf{pte\_val} \ast \ulcorner (\textsf{entry\_present pte\_val}) \urcorner \ast P \\
 \lor \; \;  \textsf{pte} \mapsto_{\textsf{vpte,1}} (\textsf{set\_pfn}(\textsf{pte\_val}, \textsf{PTE\_ADDR\_TO\_PFN}(\textsf{fpaddr}))) \ast  \ulcorner \lnot(\textsf{entry\_present pte\_val}) \urcorner \ast P \end{array} }$
;;}
\end{lstlisting}
\vspace{-1em}
  \caption{Specification and proof of code for mapping a new page with $\textsf{R}_{\textsf{walk\_acc}}$ in Figure \ref{walkpgdir} expanded.}
\label{fig:mapping_code}
\end{figure}
